---
title: "Latent Variable Measurement Error Correction with `lpmec`"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{lpmec Package Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[UTF-8]{inputenc}
---

## Introduction

This tutorial demonstrates how to use the `lpmec` package for measurement error correction in regression models using latent variable estimation. The package implements bootstrapped analyses to account for measurement error in observed indicators and provides corrected regression coefficients.

## Installation

First install the required dependencies and the `lpmec` package:
```{r setup, eval=FALSE}
# Install lpmec from source (replace with appropriate installation method)
# devtools::install_github("cjerzak/lpmec-software", subdir = "lpmec")
```


## Basic Usage

### Data Simulation

Simulate data with a latent predictor and observed binary indicators:

```{r, eval=TRUE}
set.seed(123)
n <- 1000  # Number of observations
d <- 10    # Number of observable indicators

# Generate latent variable and observed outcomes
x_true <- rnorm(n)
Yobs <- 0.4 * x_true + rnorm(n, sd = 0.35)

# Generate binary indicators of latent variable
ObservablesMat <- sapply(1:d, function(j) {
  p <- pnorm(0.5 * x_true + rnorm(n, sd = 0.5))
  rbinom(n, 1, p)
})
```

### Running the Analysis

Use lpmec to estimate corrected coefficients:

```{r, eval=TRUE}
library(lpmec)

# Run analysis with bootstrap uncertainty estimates
results <- lpmec(
  Y = Yobs,
  observables = as.data.frame(ObservablesMat),
  n_boot = 10,      # Use 0 to disable bootstrap resampling
  n_partition = 5,  # Reduced for demonstration
  estimation_method = "em"
)
```

Compare naive and corrected estimates:
```{r, eval=TRUE}
print(results)
summary(results)
```

You can visualize the relationship between split-half estimates:
```{r}
plot(results)
```

### Inference for Median Partition Aggregation

The default partition aggregation is the finite-partition median. Because the
median is nonsmooth, the formal route for uncertainty is subsampling or
m-out-of-n bootstrap with root-scaled intervals:

```{r, eval=FALSE}
results_med <- lpmec(
  Y = Yobs,
  observables = as.data.frame(ObservablesMat),
  n_boot = 499L,
  n_partition = 10L,
  estimation_method = "em",
  partition_aggregation = "median",
  bootstrap_method = "subsampling",
  boot_m_rule = "power",
  boot_m_exponent = 0.70,
  boot_ci_type = "root",
  seed = 123
)
```

This path holds the same realized partition set fixed across the original
sample and all subsamples, reruns the full latent-score and correction pipeline
inside each subsample, and rescales the empirical root statistic from `m` to
the original `n` rate. Inspect `boot_m`, `boot_m_ratio`,
`valid_partitions_by_boot`, `bootstrap_success_rate`, and sensitivity across a
grid of `boot_m` values in applied work.

## Advanced Features

### Using Different Estimation Methods

The package supports multiple estimation approaches:

```{r, eval=TRUE}
# Bayesian MCMC estimation with the default pscl backend
if(FALSE){
mcmc_results <- lpmec(
  Y = Yobs,
  observables = as.data.frame(ObservablesMat),
  estimation_method = "mcmc"
)

# To use the optional NumPyro backend, first run build_backend()
# or provide your own Python environment with JAX and NumPyro.
numpyro_results <- lpmec(
  Y = Yobs,
  observables = as.data.frame(ObservablesMat),
  estimation_method = "mcmc",
  mcmc_control = list(backend = "numpyro"),
  conda_env = "lpmec"
)
}
```

## Panel and Fixed-Effects Designs

Fixed-effects and differencing transforms strip the persistent part of a
latent signal but not its transient measurement error, so the naive panel
coefficient is attenuated far more than a pooled correction would suggest.
`lpmec_panel()` estimates the design-specific reliability of each measure,
corrects the naive OLS coefficient (`b * sqrt(rho_pooled) / rho_design`),
and reports a design-robust split-IV alternative (which multiplies by
`sqrt(rho_pooled)` instead of dividing -- IV multiplies, OLS divides).
Simulate a small panel with two noisy half scores of one latent measure:

```{r, eval=TRUE, message=FALSE}
set.seed(1234)
n_units <- 40
n_periods <- 10
unit <- rep(seq_len(n_units), times = n_periods)
time <- rep(seq_len(n_periods), each = n_units)

# Latent predictor with a persistent unit component; Y has unit fixed effects
latent_panel <- rnorm(n_units)[unit] + rnorm(n_units * n_periods, sd = 0.7)
half1 <- latent_panel + rnorm(n_units * n_periods, sd = 0.5)
half2 <- latent_panel + rnorm(n_units * n_periods, sd = 0.5)
Y_panel <- 0.4 * latent_panel + rnorm(n_units)[unit] +
  rnorm(n_units * n_periods)

panel_results <- lpmec_panel(
  Y = Y_panel, unit = unit, time = time,
  split_scores = list(measure1 = cbind(half1, half2)),
  design = "within",
  n_boot = 3,   # Increase in applied work
  seed = 42
)
```

The summary shows the naive within-unit OLS estimate next to the corrected
OLS coefficient (with its reliability-variant interval) and the corrected
split-IV estimate, with cluster-bootstrap standard errors:

```{r, eval=TRUE}
print(panel_results)
summary(panel_results)
```

The half scores used for the reliability estimates can be inspected
visually:

```{r, eval=TRUE}
plot(panel_results)
```

## Reliability Bounds with Multiple Measures

With three or more measures of the same latent trait,
`lpmec_reliability_bounds()` brackets each measure's reliability between the
within-measure split correlation (an upper candidate when halves share
method error) and the cross-measure triad estimate
`rho*_m = r_ml * r_mk / r_lk` (a lower candidate under cross-measure error
dependence), reporting the identification interval `[rho_lo, rho_hi]`:

```{r, eval=TRUE, message=FALSE}
set.seed(2025)
n_rb <- 400
latent_rb <- rnorm(n_rb)
rb_half1 <- latent_rb + rnorm(n_rb, sd = 0.6)
rb_half2 <- latent_rb + rnorm(n_rb, sd = 0.6)
score2 <- latent_rb + rnorm(n_rb, sd = 0.7)
score3 <- latent_rb + rnorm(n_rb, sd = 0.7)

bounds <- lpmec_reliability_bounds(
  split_scores = list(measure1 = cbind(rb_half1, rb_half2)),
  scores = list(measure2 = score2, measure3 = score3)
)
print(bounds)
```

Passing `unit`, `time`, and `designs = c("pooled", "within")` reports how
much each measure's reliability deteriorates under the design transform
actually used for estimation.

## Latent Moderators in Experiments

When a randomized treatment's effect is moderated by a latent trait measured
with error, the naive interaction coefficient from `Y ~ treatment * x`
converges to `b_TX * sqrt(rho)` for a standardized moderator score of
reliability `rho`. `lpmec_moderator()` estimates the Spearman-Brown as-used
reliability of the score entered in the regression and divides the naive
interaction by its square root. Simulate an experiment with a 4-item
moderator battery:

```{r, eval=TRUE, message=FALSE}
set.seed(321)
n_mod <- 400
X_mod <- rnorm(n_mod)
treatment <- rbinom(n_mod, 1, 0.5)
Y_mod <- 0.2 * treatment + 0.2 * X_mod +
  0.3 * treatment * X_mod + rnorm(n_mod)
items <- sapply(c(1.2, 1.0, 0.8, 1.1), function(discrimination) {
  rbinom(n_mod, 1, pnorm(discrimination * X_mod - 0.2))
})
colnames(items) <- paste0("item", 1:4)

moderator_results <- lpmec_moderator(
  Y = Y_mod,
  treatment = treatment,
  observables = items,
  estimation_method = "averaging",
  n_boot = 5,       # Increase in applied work
  n_partition = 2,  # Reduced for demonstration
  seed = 99
)
```

The corrected interaction is reported with bootstrap intervals that
propagate the uncertainty in the reliability estimate itself, and
`rho_score` appears as a headline quantity:

```{r, eval=TRUE}
print(moderator_results)
summary(moderator_results)
```

## Conclusion

The `lpmec` package provides robust measurement error correction through:

  - Bootstrapped latent variable estimation
  - Multiple correction methods (OLS, IV, Bayesian)
  - Flexible estimation backends (EM, MCMC)
  - Panel/fixed-effects and differencing-design corrections (`lpmec_panel`)
  - Reliability bounds from split and triad estimates (`lpmec_reliability_bounds`)
  - Latent-moderator interaction corrections in experiments (`lpmec_moderator`)

Refer to package documentation for advanced configuration options.
